---
title: "Setting Objective Function"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{objective}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(easylp)
```

# Introduction

This vignette will explore the different ways to define and interpret the objective function.

## Basics

```{r}
lp <- easylp$new()
lp$var("x", myset = letters[1:5])
```

# Posterior Transformations

In a linear problem, the objective function must always be a linear function. Internally, this means each variable has a coefficient. Say we have an objective function such as this one:

$$
\min{z = \log(x+y)}
$$

This is not a linear function, but it can be converted to an equivalent one. Because the *log* function is [monotonically increasing](https://en.wikipedia.org/wiki/Monotonic_function), minimizing that problem is the same as minimizing this one.

$$
\{\min{z = \log(x+y)}\} \equiv \{\min{z' = x+y}\}
$$

This is to say, the optimal solution will be the same, but not the objective value.
