---
title: "Setting Objective Function"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setting Objective Function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(easylp)
```

# Introduction

This vignette will explore the different ways to define the objective function. It does not assume any prior knowledge of the package. However, it's recommended to start with `vignette("easylp")`.

## Basics

The objective function is an expression that establishes the goal of the problem, as well as the role of each variable in achieving it. For example, let's say we're a company maximizing profits. We would define the objective function to be something like this.

```{r}
lp <- easylp$new()
lp$var("earnings")
lp$var("costs")
lp$max(earnings - costs)
```

If you have worked with linear programming before, you will know that the objective function is simply a vector of coefficients for each variable. We see this vector using `$objective_fun`.

```{r}
lp$objective_fun
```

After defining some constraints and solving the problem, we can get the objective value, our profits, by using `$objective_value`.

```{r}
lp$con(
    earnings <= 3*costs,
    costs <= 50
)

lp$solve()
lp$objective_value
```

The objective value is also visible in the summary of the problem.

```{r}
print(lp)
```

## Direction

The direction of a problem represents whether the goal is to minimize or maximize the objective function. In EasyLP, it's set by using `$min()` or `$max()` when defining the objective.

```{r}
lp <- easylp$new()
lp$var("x")
lp$min(x); lp$direction
lp$max(x); lp$direction
```

It can also be changed by assigning `"min"` or `"max"` to the field. This is not recommended, as it makes the problem less readable.

```{r}
lp$min(x)
lp$direction <- "max"
lp$direction
```

# Defining the Objective Function

The result of the objective function must be a single value. For instance, the following code will throw an error.

```{r, error=TRUE}
lp <- easylp$new()
lp$var('x', my_set = 1:3)
lp$min(x)
```

The message tells us to wrap the values in a `sum()` such that the 3 values become 1. This section discusses different methods of wrapping values.

## Sum and Mean

These functions work just like they do in R. Sum adds all the values from an array, and mean divides this result by the length of the array. Weighted means are also supported.

```{r}
lp <- easylp$new()
lp$var('x', rows = 1:2, cols = LETTERS[1:3])
lp$var('y', rows = 1:2)
( lp$min(sum(x)) )
( lp$min(sum(x, y)) )
( lp$min(sum(x) + mean(y)) )
( lp$min(sum(x) + weighted.mean(y, c(1, 3))) )
```

## Sums and Means Across Margins

Functions `rowSums(), colSums(), rowMeans(), colMeans()` are supported, as well as `apply()`. For more information about these functions, see their documentation in base R. Generally, their use is not recommended in favor of `sum_for()`.

```{r}
( lp$min(sum(colSums(x) * 1:3)) )
( lp$min(sum(rowSums(x) * 4:5)) )
```

### Note for Advanced Users

These functions are not methods, like `sum()` or `mean()`. Instead, they are redefined in a hidden environment, where the constraints and objective function are evaluated.

## Sum For

The `sum_for()` notation is similar to the way sums are written in math. See these 2 examples to understand how to use it, and read the documentation for more information.

$$
\min\sum_{i=1}^{2}{y_i \cdot (a_i + i)}
$$

```{r}
a = 4:5
(lp$min(
    sum_for(i=1:2, y[i] * (a[i] + i))
))
```

$$
\max\sum_{i=1}^2 \sum_{j\in\{A,B,C\}} {(x_{ij} + y_i) \cdot b_j}
$$

```{r}
b = c(A=3, B=1, C=2)
(lp$min(
    sum_for(i=1:2, j=names(b), (x[i,j] + y[i]) * b[j])
))
```

# Posterior Transformations

In a linear problem, the objective function must always be a linear function. Internally, this means each variable has a coefficient.

## Addend

EasyLP supports an addend, an extra coefficient to add or subtract from the objective value.

Say that a company has a fix cost of 50, that doesn't depend on the production. We could implement it this way

```{r}
lp <- easylp$new()
lp$var("x", lower=10)
lp$var("y", lower=10)
lp$min(4*x + 3*y + 50)
lp$solve()
lp
```

Notice how it's printed: 70 is the result of the internal objective function $4x + 3y$. Then, EasyLP adds the fixed cost of 50.

If we call `lp$objective_value`, it returns 120. We can get the *raw* objective value (before transformation) by calling `lp$objective_value_raw`. Usually, these two are the same: they only differ when a posterior transformation is applied.

```{r}
lp$objective_value_raw
lp$objective_add
lp$objective_value
```

## Non-Linear Transformations

Say now we have an objective function such as this one:

$$
\min{z = \log(x+y)}
$$

This is not a linear function, but it can be converted to an equivalent one. Because the *log* function is [monotonically increasing](https://en.wikipedia.org/wiki/Monotonic_function), minimizing that problem is the same as minimizing this one.

$$
\{\min{z = \log(x+y)}\} \equiv \{\min{z' = x+y}\}
$$

This is to say, the optimal solution will be the same, but the objective value will not. It is not possible to use a logarithm in EasyLP and it will throw an error, telling us that function *log* is not linear.

```{r, error=TRUE}
lp$min(log(x + y))
```

Instead, we can use the `transform` argument to add a transformation.

```{r}
lp$min(x + y, transform = log)
lp$solve()
lp
```

The `transform` argument supports lambda notation. All of the following work:

```{r}
lp$min(x + y, log)
lp$min(x + y, ~log(.))
lp$min(x + y, \(obj) log(obj))
lp$objective_transform <- function(obj) {
    log(obj)
}
```

### Warning!

Transformations should be [monotonically increasing](https://en.wikipedia.org/wiki/Monotonic_function) functions. Otherwise, the problem is not equivalent. EasyLP does not currently check if functions satisfy this condition. If a function is monotonically decreasing, you should invert the direction of the problem, or redefine the function such that it increases. Keep in mind the solver optimizes the linear function, then transforms it.

### Note

The transformation is applied every time `lp$objective_value` is called. To use the value, consider assigning it to a variable to avoid calculating it every time.

## Both Addend and Posterior

It's possible to use both an addend and a posterior transformation, but beware! It is first added then transformed.

`objective_value == transform(raw_value + addend) != transform(raw_value) + addend`

```{r}
lp$min(x + y + 10, ~ .*2)
lp$solve()
lp
```
